---
title: "College Degrees And Income"
author: "Hanh Nguyen"
date: "12/10/2018"
output: 
  html_document:
    df_print: kable

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
degrees <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2018-10-16/recent-grads.csv")
```

#Overview

Data was obtained from SourceGithub -> tidytuesday -> recent-grads. There are `r ncol(degrees)` variables and `r nrow(degrees)` observations in the dataset.

The analysis is based on an article about the relationship between Degrees and Income. The issue we need to keep in mind is that the dataset is from 2010. 

Data types for each variable are below.

```{r,echo= FALSE}
kable(sapply(degrees,class))
```

Initial Questions:

* What major category is most/least popular?
* Major category difference by gender?
* What major is struggling the most to find a job? What major is easier to find a job? 
* Do women struggle more than men to find employment, agnostic of major
* Relationship between earnings, gender gap, and unemployment rate.

## 1. ENROLLMENT DATA

### 1.1 What major category is most/least popular?

Across 16 College Majors, Business is the most popular college major category with highest level of enrollment at\ 1,302,376.

```{r}
degrees %>%
  group_by(Major_category) %>%
  filter(!is.na(Total)) %>%
  summarize(total_enrollment = sum(Total)) %>%
  ggplot(aes(x= reorder(Major_category,total_enrollment),y=total_enrollment, fill = Major_category))+ geom_col() + coord_flip() + labs(x = "College Majors", y = "Total Enrollment")
```

### 1.2 Men/Women are really different in choosing major?

College Majors with the largest gender gaps. 

```{r}
by_major_category <- degrees %>%
  filter(!is.na(Total)) %>%
  group_by(Major_category) %>%
  summarize(Men = sum(Men),
            Women = sum(Women),
            Total = sum(Total))%>%
   mutate(Share_Women = Women / Total)%>%
  filter(Share_Women >= 0.6 | Share_Women <= 0.4)

by_major_category %>%
  arrange(desc(Total))%>%
  select(Major_category, Men, Women, Total) %>%
  gather(key = "Gender", value = "Enrollment", Men, Women)%>%
  ggplot(aes(x = reorder(Major_category, Enrollment), y = Enrollment, fill = Gender)) + geom_col(position ="dodge") + coord_flip() + labs(x = "College Majors", y = "Total Enrollment") 
```

## 2. EMPLOYMENT DATA

### 2.1  Employability by major?

No clear pattern between Major_category and Unemployment_rate. 

```{r}
degrees %>%
  filter(!is.na(Total))%>%
  group_by(Major_category) %>%
  summarize(un_rate = mean(Unemployment_rate)*100, total_enrollment = sum(Total))%>%
  ggplot(aes(x= reorder(Major_category, total_enrollment),y=un_rate, fill = Major_category))+ geom_col() + coord_flip() + labs(x = "College Majors", y = "Unemployment rate (%)") 
```

**But don't assume all college sub majors are the same!**. We dive deeper to see what *college sub majors* students struggle most to find a job and what *college sub majors* where it is easier to find a job. 

*Top 5 sub majors, highest rate of unemployment*

Surprisingly, the major with highest rate of unemployment- Nuclear Engineering belongs to Engineering with nearly 20% unemployed. 
```{r}
degrees %>%
  arrange(desc(Unemployment_rate))%>%
  head(5)%>%
  select(Major_category, Major, Total, Unemployment_rate)%>%

knitr::kable()
```

*Top 5 sub majors, lowest rate of unemployment*

Bottom line, there are wide differences among majors within a major category. For example, an engineer can still have high chance of being unemployed.  

```{r}
degrees %>%
  arrange(desc(Unemployment_rate))%>%
  tail(5)%>%
  select(Major_category, Major, Total, Unemployment_rate)%>%

knitr::kable()
```

### 2.2 Full time job and part time job by college majors, ranked by enrollment

One observation is that the College major with the highest enrollment (Business) also has the highest number in Full time employment.

```{r}
degreesE <- degrees%>%
  filter(!is.na(Total)) %>%
  group_by(Major_category)%>%
  summarize(total_enrollment = sum(Total))

degrees %>%
  group_by(Major_category)%>%
  summarize(full_time = sum(Full_time), part_time = sum(Part_time))%>%
  gather(key = "EmploymentType", value = "Employed", full_time, part_time) %>%
  left_join(degreesE, by = "Major_category") %>%
  ggplot(aes(x = reorder(Major_category, total_enrollment), y = Employed, fill = EmploymentType)) + geom_col(position ="stack") + coord_flip() + labs(x = "College Majors", y = "Employment") 
```

### 2.3 College job and non college job by college majors, ranked by enrollment

But business is also the college major which has the highest percentage of employment in a non-college job (that means a job that doesn't require that degree).

```{r}
degreesE <- degrees%>%
  filter(!is.na(Total)) %>%
  group_by(Major_category)%>%
  summarize(total_enrollment = sum(Total))

degrees %>%
  group_by(Major_category)%>%
  summarize(college = sum(College_jobs), non_college = sum(Non_college_jobs))%>%
  gather(key = "EmploymentType", value = "Employed", college, non_college) %>%
  left_join(degreesE, by = "Major_category") %>%
  ggplot(aes(x = reorder(Major_category, total_enrollment), y = Employed, fill = EmploymentType)) + geom_col(position ="stack") + coord_flip() + labs(x = "College Majors", y = "Employment") 
```

### 2.4 Unemployment rate by gender and Major

* Male dominated sub-majors: sub majors that have percentage of Women signing up less than 50%
* Female dominated sub-majors: sub majors that have percentage of Women signing up more than 50%

Both graphs exhibit similar pattern. Most male dominated sub-majors tend to have unemployment rate around 4%-8% and then 9-10%  (we see a more stacked cluster in those areas and then thinner in other areas). Female dominated sub-majors tend to have unemployment rate from 5-8% and mostly less than 10%. 

Because the size of the circle represents the median income, we can see that male-dominated graph has more big circle and the scale of the circle also increases by \$10,000 while female-dominated graph has fewer big circle and the scale of the circle also increass by less, just \$5,000. 

```{r}
degrees_3 <- degrees %>%
    filter(ShareWomen <= .5) %>%
    arrange(desc(Men)) %>% 
    head(50)
  
 ggplot(degrees_3, aes(x = Men, y = Unemployment_rate, size = Median, color = Major_category)) + 
    geom_point() + ggtitle("Unemployment rate for sub-majors dominated by men")
```

```{r}
degrees_2 <- degrees %>%
  filter(ShareWomen > .5) %>%
  arrange(desc(Women)) %>% 
  head(50)

  ggplot(degrees_2, aes(x = Women, y = Unemployment_rate, size = Median, color = Major_category)) + 
  geom_point() + ggtitle("Unemployment rate for sub-majors dominated by women")
```

## 3. EARNINGS DATA

## 3.1 Median Income by College Majors

* Engineering has highest Median income.
* Out of all sub-majors, Petroleum Engineering students have the highest Median income after graduation.
* Out of all sub-majors, Library Science student have the lowest Median income after graduation. 

```{r}
outliers <- degrees%>%
  mutate(rank = rank(desc(Median))) %>%
  filter(rank == 1 | rank == 173)
  
degrees %>%
  group_by(Major_category)%>%
  ggplot(aes(x = reorder(Major_category, Median, FUN = median), y = Median)) + geom_boxplot() + coord_flip() + labs(x = "Major Category", y = "Median Income") + geom_text(data = outliers, aes(label = Major), checkoverlap = TRUE, vjust = "inward", hjust = "inward") 
```

### 3.2 Top 10 Majors with highest earnings, ranked by enrollment

8/10 sub Majors belong to Engineering, only 1 in Business (Actuarial Science) and 1 in Physcial Sciences.
Interestingly, Nuclear Engineering has the highest unemployment rate, yet 4th in terms of enrollment and also similar rank in terms of earnings. 

```{r}
degrees%>%
  group_by(Major)%>%
  arrange(desc(Median))%>%
  head(10)%>%
  ggplot(aes(x= reorder(Major, Total), y = Median, color = Major_category)) + geom_point() + geom_errorbar(aes(ymin =P25th, ymax= P75th)) + coord_flip() + labs(x = "Sub Majors", y = "Median income") + scale_y_log10() 
```

## 4. GENDER AND EARNINGS

### 4.1 First, we build a linear regression model to see how variation in Median Income is explained by multiple independent variables.

```{r}
model <- lm(Median ~ ShareWomen + Low_wage_jobs + Full_time + Unemployment_rate, data = degrees)
summary(model)
```

### 4.2 A further look at ShareWomen

As Percentage of women increases, the median income decreases.

*P value is <2 e-16, which means that percentage of women significantly represent median income.
*The coefficient is -2.860e+4 which means they are negative related.

It doesn't necessarily mean that more women in that major, lower earnings for that major. We just witness a negative correlation and that doesn't indicate a causal relationship. There are a lot of other factors involved. 

```{r}
degrees %>%
ggplot(aes(x = ShareWomen, y = Median, color = Major_category)) + geom_point() + geom_smooth(aes(color = NULL), method = "lm") + labs(y = "Median_income", x = "Percentage of Women")
```

### 4.3 Gender gap earnings, break down by college majors

What we see in this interesting graph is within a major category, how male-dominated major is paid compared to female dominated major? It can be seen that it's not always male-dominated major is paid higher than female-dominated major. 

```{r}
degrees$Type <- NA
for (i in 1:nrow(degrees)){
  if(is.na(degrees$ShareWomen[i]) || degrees$ShareWomen[i] >= 0.5){
    degrees$Type[i] <- "Women dominate"
  } else if(is.na(degrees$ShareWomen[i]) || degrees$ShareWomen[i] < 0.5){
    degrees$Type[i] <- "Men dominate"
  }
}
degrees %>% 
  filter(!is.na(Total)) %>%
  group_by(Major_category)%>%
  ggplot(aes(x= reorder(Major_category, Median, FUN = median), y = Median, fill = Type)) + geom_boxplot() + coord_flip() + 
  labs(x = "Major Category", y = "Median Income") + scale_y_log10()
```

### 4.4 A further look at unemployment rate

Unemployment rate does not significantly represent median income

* P value is 0.5 which is much larger than 0.05 level.
* The coefficient is -1.465e+4.

```{r}
degrees %>%
ggplot(aes(x = Unemployment_rate, y = Median, color = Major_category)) + geom_point() + geom_smooth(aes(color = NULL), method = "lm")
```

## 5. CONCLUSION

Some key take-aways from the project:

* Business Major is most popular but Engineer Major earns the most, in terms of median income.
* Not all sub majors will be yield the same result, even under the same college major. 
* Business generally has the highest level of full time employment relative to part time employment. However, business also has the highest percentage of non-college job relative to college jobs.
* Although we set out to see if gender gap can affect the earnings, we can't really conclude a causal, at least with this dataset. All we see is that there is a negative correlation between gender and earnings.
* The reason for this is college majors: women tend to choose a major that pays lower and men tend to choose a major that pays higher.
* But that doesn't hold true all the time: in some college majors, female-dominated sub-majors pay significantly higher than male-dominated sub-majors, so gender gap is not an issue here. 
* Next step could be a more thorough investigation of geography, location of work, etc to get a bigger picture. The dataset is limited to generate any meaningful analysis and recommendation. 
* However, there still be a big call-out across the analysis:

### STAY AT SCHOOL, KID!!!

